Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization

### 2023-10-30 manuscript.Rmd
cite deviation paper
cite fiedler again

One widely known factor is the contribution of statistical methods to theoretical risk.
Theoretical risk $P(\neg{}E|\neg{}H)$ is deeply connected with statistical methods, because it is related to the type-I error rate in statistical hypothesis testing $P(E|\neg{}H)$ by $P(\neg{}E|\neg{}H) = 1 - P(E|\neg{}H)$, if you consider the overly simplistic case where the research hypothesis is equal to the statistical alternative-hypothesis because then the nill-hypothesis is $\neg{}H$.
Because many researchers are familiar with the type-I error rate, it can be helpful to remember this connection to theoretical risk.
Researchers who choose a smaller type-I error rate can be more sure of their results, if significant, because the theoretical risk is higher.
However, this connection should not be overinterpreted for two reasons.
First, according to most interpretations of null hypothesis testing, the absence of a significant result should not generally be interpreted as evidence against the hypothesis [@mayoStatisticalInferenceSevere2018, 5.3].
Second, the research hypothesis rarely equals the statistical alternative hypothesis (most research hypothesis are more specific than "any value except zero").
In fact, it is entirely possible to assume the null hypothesis as a research hypothesis, as is commonly done in e.g., structural equation modelling, where the roles of detectability, theoretical risk and type-I/II error rate switch.
We argue that theoretical risk (and hence its complement, $P(E|\neg{}H)$) also encompasses factors outside the statistical realm, most notably the study design and broader analytical strategies.
Type-I error rate is the property of a statistical test under some assumptions, whereas theoretical risk is a researchers' belief.
One may take such theoretical properties as a first starting point to form a substantive belief but surely researchers ought to take other factors into consideration.
For example, if a researcher believes that there might be confounding variables at play for the relation between two variables, this should decrease theorethical risk; after all they might find an association purely on account of the confounders.

@@

Second, researchers often have to deviate from the preregistration and make data-dependent decisions after the preregistration.
If the only goal of preregistration is to ensure confirmatory research, such changes are not justifiable.
However, under our rational, some changes may be justified.
Any change increases the uncertainty about the theoretical risk and may even decrease the theoretical risk.
The changes still may be worthwhile if the negative outcomes may be offset by an increase in detectability due to the change.
Consider a preregistration that failed to specify how to handle missing values, and researchers subsequently encountering missing values.
In such case, detectability becomes zero because the data cannot be analyzed without a post-hoc decision about how to handle the missing data.
Any such decision would constitute a deviation from the preregistration, which is possible under our proposed objective.
Note that a reader cannot rule out that the researchers leveraged the decision to decrease theoretical risk, i.e., picking among all options the one that delivers the most beneficial results for the theory (in the previous example, chosing between various options of handling missing values).
Whatever decision they make, increased uncertainty about the theoretical risk is inevitable and the expected epistemic value is decreased compared to a world where they anticipated the need to deal with missing data.
However, it is still justified to deviate.
After all they have not anticipated the case and are left with a detectablilty of zero.
Any decision will increase detectability to a non-zero value offsetting the increase in uncertainty.
The researchers also may do their best to argue that the deviation was not motivated by increasing theoretical risk, thereby, decreasing the uncertainty.
Ideally, there is a default decision that fits well with the theory or with the study design.
Or, if there is no obvious candidate, the researchers could conduct a multiverse analysis of the available options to deal with missings to show the influence of the decision [@steegenIncreasingTransparencyMultiverse2016].

>>>>

One widely known factor is the contribution of statistical methods to theoretical risk.
Theoretical risk $P(\neg{}E|\neg{}H)$ is deeply connected with statistical methods, because it is related to the type-I error rate in statistical hypothesis testing $P(E|\neg{}H)$ by $P(\neg{}E|\neg{}H) = 1 - P(E|\neg{}H)$, if you consider the overly simplistic case where the research hypothesis is equal to the statistical alternative-hypothesis because then the nill-hypothesis is $\neg{}H$.
Because many researchers are familiar with the type-I error rate, it can be helpful to remember this connection to theoretical risk.
Researchers who choose a smaller type-I error rate can be more sure of their results, if significant, because the theoretical risk is higher.
However, this connection should not be overinterpreted for two reasons.
First, according to most interpretations of null hypothesis testing, the absence of a significant result should not generally be interpreted as evidence against the hypothesis [@mayoStatisticalInferenceSevere2018, 5.3].
Second, the research hypothesis rarely equals the statistical alternative hypothesis (most research hypothesis are more specific than "any value except zero").
In fact, it is entirely possible to assume the null hypothesis as a research hypothesis, as is commonly done in e.g., structural equation modelling, where the roles of detectability, theoretical risk and type-I/II error rate switch.
We argue that theoretical risk (and hence its complement, $P(E|\neg{}H)$) also encompasses factors outside the statistical realm, most notably the study design and broader analytical strategies.
Type-I error rate is the property of a statistical test under some assumptions, whereas theoretical risk is a researchers' belief.
One may take such theoretical properties as a first starting point to form a substantive belief but surely researchers ought to take other factors into consideration.
For example, if a researcher believes that there might be confounding variables at play for the relation between two variables, this should decrease theoretical risk; after all they might find an association purely on account of the confounders [@fiedlerWhatConstitutesStrong2017].

@@

Second, researchers often have to deviate from the preregistration and make data-dependent decisions after the preregistration.
If the only goal of preregistration is to ensure confirmatory research, such changes are not justifiable.
However, under our rational, some changes may be justified.
Any change increases the uncertainty about the theoretical risk and may even decrease the theoretical risk.
The changes still may be worthwhile if the negative outcomes may be offset by an increase in detectability due to the change.
Consider a preregistration that failed to specify how to handle missing values, and researchers subsequently encountering missing values.
In such case, detectability becomes zero because the data cannot be analyzed without a post-hoc decision about how to handle the missing data.
Any such decision would constitute a deviation from the preregistration, which is possible under our proposed objective.
Note that a reader cannot rule out that the researchers leveraged the decision to decrease theoretical risk, i.e., picking among all options the one that delivers the most beneficial results for the theory (in the previous example, chosing between various options of handling missing values).
Whatever decision they make, increased uncertainty about the theoretical risk is inevitable and the expected epistemic value is decreased compared to a world where they anticipated the need to deal with missing data.
However, it is still justified to deviate.
After all they have not anticipated the case and are left with a detectablilty of zero.
Any decision will increase detectability to a non-zero value offsetting the increase in uncertainty.
The researchers also may do their best to argue that the deviation was not motivated by increasing theoretical risk, thereby, decreasing the uncertainty.
Ideally, there is a default decision that fits well with the theory or with the study design.
Or, if there is no obvious candidate, the researchers could conduct a multiverse analysis of the available options to deal with missings to show the influence of the decision [@steegenIncreasingTransparencyMultiverse2016].
In any case, deviations must be transparently reported and we applaud recent developments to standardize and normalize this process [@willrothBestLaidPlans2023].
